\pdftitle

### 2018-05-14 maxNt_v2.tex
V2

%% Author: PGL  Porta Mana
%% Created: 2015-11-04T09:20:08+0100
%% Last-Updated: 2018-05-13T22:11:43+0200
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%Report-no: ***
\documentclass{article}

@@

Some investigations in the literature proceed as follows. We bin in time
and binarize in intensity \cite{caianiello1961,caianiello1986} the recorded
activity, transforming it into a sequence of $0$s (inactive) and $1$s
(firing) for each neuron. Groups of two or more $1$s can appear in any time
bin. Then we ask a question of this kind: can the presence of groups of
three or more simultaneous firings be \enquote{explained} in terms of
groups of just two, or three, or\ldots? This kind of question is very vague
until we define what we mean with \enquote{explain} or similar words.
Usually such \enquote{explanation} has a probabilistic meaning: we wonder
whether the sole knowledge about two, or three, or more simultaneous
firings can yield as good predictions about the whole activity as a more
complete knowledge would do. To make this question concrete we must choose
a probabilistic model.

Maximum-entropy models
\cite{jaynes1957,jaynes1963,sivia1996_r2006,meadetal1984} have been used in
neuroscience with this kind of purpose and for diverse brain areas at least
since the 1990s
\cite{mackay1991,martignonetal1995,bohteetal2000,amarietal2003,schneidmanetal2006,shlensetal2006,mackeetal2009b,roudietal2009c,tkaciketal2009,gerwinnetal2009,mackeetal2011,mackeetal2011b,ganmoretal2011,granotatedgietal2013,tkaciketal2014b,moraetal2015,shimazakietal2015}.
Part of this literature focused on the range of applicability of \me\
distributions
\cite{tkaciketal2006,roudietal2009,roudietal2009b,barreiroetal2010b,mackeetal2013,rostamietal2016_r2017}.
We assume that our readers are at least vaguely familiar with such models.
In brief, they build a probability distribution for neuronal activity using
only data about simultaneous firings of two neurons, or of three, and so
on. Maximum-entropy probability distributions can be introduced and justified in
three main ways:
\begin{enumerate}
\item from the maximum-entropy method
  \cite{jaynes1957,jaynes1963,sivia1996_r2006};
\item as asymptotic distributions of Bayesian models called
  \enquote{exchangeable} with entropic priors
  \cites{jaynes1986d_r1996,skilling1989b,skilling1990,rodriguez1989,rodriguez2002,portamana2017}[see
  also][\chap~VI]{schroedinger1946_r1948}%[pp.~52--55]
{jaynes1979b};
\item as asymptotic distributions of Bayesian models based on sufficient
  statistics
  \cite{jaynes1982,%[\sect~I]
jaynes1982b,portamana2017}.
\end{enumerate}

*** usually \enquote{explain} has a probabilistic meaning. We want to see
whether information about moments give reasonable predictions about the
rest. Necessity of making this precise. Three ways:

This kind of studies would have little significance if we did not expect
their results to generalize beyond the particular recorded sample. In other
words, we implicitly or explicitly expect that the conclusions drawn with
\me\ methods about the recorded neurons should somehow generalize or
extrapolate at least to whole areas those neurons are recorded from, \eg\
retina or motor cortex. We are thus assuming that the recorded sample is a
\emph{representative sample} of that area. Whether such extrapolation makes
biological sense is a difficult question, to be analysed case by case.

@@

But a problem remains even if we assume that the sample is representative
of its brain area. The problem lies in the way the \me\ methods
is typically used in the literature. In this note we want to discuss this
problem and offer a solution to it.

@@

>>>>

%% Author: PGL  Porta Mana
%% Created: 2015-11-04T09:20:08+0100
%% Last-Updated: 2018-05-14T14:56:53+0200
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%Report-no: ***
\documentclass{article}

@@

Some investigations in the neuroscientific literature proceed as follows.
We bin in time and binarize in intensity
\cite{caianiello1961,caianiello1986} the recorded activity, transforming it
into a sequence of $0$s (inactive) and $1$s (firing) for each neuron.
Groups of two or more $1$s can appear in any time bin. Then we ask a
question of this kind: can the presence of groups of three or more
simultaneous firings be \enquote{explained} in terms of groups of just two,
or three, or\ldots? This question is very vague until we define what we
mean with \enquote{explain} or similar words. Usually such
\enquote{explanation} has a probabilistic meaning: we wonder whether the
sole knowledge about two, or three, or more simultaneous firings can yield
as good predictions about the whole activity as a more complete knowledge
would do. To make this question concrete we must first choose a
probabilistic model from some well-defined assumptions, and then assess
whether its final probabilities are \enquote{good} according to some
well-defined measure. In the following we consider only the first,
model-building, step.

Maximum-entropy probability models
\cite{jaynes1957,jaynes1963,sivia1996_r2006,meadetal1984} have been used in
neuroscience with this kind of purpose, for recordings from brain areas as
diverse as retina and motor cortex, at least since the 1990s
\cite{mackay1991,martignonetal1995,bohteetal2000,amarietal2003,schneidmanetal2006,shlensetal2006,mackeetal2009b,roudietal2009c,tkaciketal2009,gerwinnetal2009,mackeetal2011,mackeetal2011b,ganmoretal2011,granotatedgietal2013,tkaciketal2014b,moraetal2015,shimazakietal2015}.
Part of this literature focuses on the range of applicability of \me\
distributions
\cite{tkaciketal2006,roudietal2009,roudietal2009b,barreiroetal2010b,mackeetal2013,rostamietal2016_r2017}.
We assume that our readers are at least vaguely familiar with this kind of model.
In brief, it builds a probability distribution for neuronal activity using
only data about simultaneous firings of two neurons, or of three, and so
on. Let us take the case of the first two moments as an example. The
maximum-entropy distribution can be derived in three main ways:
\begin{enumerate}
\item From the maximum-entropy method
  \cite{jaynes1957,jaynes1963,sivia1996_r2006}. The distribution is chosen
  as the one with the highest Shannon entropy among those having particular
  first and second moments. This distribution is deemed to be
  \enquote{maximally noncommittal} \cite{jaynes1963}, but this adjective
  means little without further technical characterization. Two
  characterizations are given by the next two derivations.

\item As the asymptotic distribution of a Bayesian model called
  \enquote{exchangeable}, with a particular prior
  \cites{jaynes1986d_r1996,skilling1989b,skilling1990,rodriguez1989,rodriguez2002,portamana2017}[see
  also:][\chap~VI]{schroedinger1946_r1948}%[pp.~52--55]
  {jaynes1979b}. This distribution is the limit, when the number of time
  bins is very large, of
  \begin{equation}
    \label{eq:predictive_exchangeability}
    \p(\text{\small activity in one time bin} \|
    \text{\small 1st \amp\ 2nd empirical moments from other bins}%,\text{\small other info}
    ),
  \end{equation}
  calculated with Bayes's theorem from a particular initial joint
  distribution in which different time bins are almost independent. This
  derivation produces a probability just by taking into account those two
  moments, but does not exclude that other information might be relevant.
  It can therefore be compared with a probability that uses additional
  information, or with empirical frequencies.
\item As the asymptotic distribution of a Bayesian model based on
  sufficient statistics \cite{jaynes1982,%[\sect~I]
    jaynes1982b,portamana2017}. This distribution is the limit, when the
  number of time bins is very large, of
\begin{equation}
  \label{eq:general_prediction}
  \p(\text{\small activity in one time bin} \|
  \text{\small activities in other bins}%,\text{\small other info}
  )
\end{equation}
under the assumption that \emph{only} the values of the first and second
empirical moments of the activities are relevant for the probability above,
\ie\
\begin{multline}
  \label{eq:informal_sufficiency}
  \p(\text{\small activity in one time bin} \|
  \text{\small activities in other bins}%,\text{\small other info}
  ) \equiv{}\\
  \p(\text{\small activity in one time bin} \|
  \text{\small 1st \amp\ 2nd empirical moments from other bins}%,\text{\small other info}
  );
\end{multline}
in other words, the assumption that the first and second moments are
sufficient statistics. This limit generally does not depend on the initial
joint distribution. This derivation produces a probability by explicitly
assuming that only the first and second moments are informationally
relevant: any other information would lead to the same conditional
probability. It can therefore only be compared with empirical frequencies.
\end{enumerate}
We refer our readers to the cited references for a deeper discussion of
these derivations. It is remarkable that all three lead to the same
distribution. We shall now see, however, that if we apply these derivations
more carefully they lead to different distributions.

@@

This kind of investigation of recorded samples of neuronal activity would
have little significance if we did not expect its results to generalize
beyond the particular recorded sample. In other words, we implicitly or
explicitly expect that the conclusions drawn with \me\ methods about the
recorded neurons should somehow generalize or extrapolate at least to whole
areas those neurons are recorded from, \eg\ retina or motor cortex. We are
thus assuming that the recorded sample is a \emph{representative sample} of
that area. Whether such extrapolation makes biological sense is a difficult
question, to be analysed case by case. For the moment let's assume that it
does make biological sense, and that our sample can be assumed to be
representative of its brain area.

Then a problem appears in the way the \me\ method is typically applied, in
the literature, to recorded or artificial samples. In this note we want to
discuss this problem and offer a solution to it.

@@

----------- cut text ---------------------

usually \enquote{explain} has a probabilistic meaning. We want to see
whether information about moments give reasonable predictions about the
rest. Necessity of making this precise. Three ways:
